A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-11-09 jour/jour.tex
minor
tuning
more
Rearrange, describe figures.

Before constructing a model for $F\!$, the theory solver constructs neither $\mu$-terms nor the mapping $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To detect whether \rn{Acyclic} applies,
the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all constructor arguments in these equivalence classes recursively.
If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $\SSS$ of all codatatype equivalence classes. It
simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of ordinary types.
It then partitions each set %in this partition %in the same manner
recursively.
If %after doing so,
the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
and $v$, it deduces that \rn{Unique} applies to $u$ and~$v$.

@@

\begin{figure}[t!]
\normalsize
\centering
\begin{enumerate}
\item Let $\mathcal{U}$ be a partition of the equivalence classes of $\Ec$ such that $\ec{t} =_{\mathcal{U}} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if:
\begin{itemize}
\item $t\typ{\delta}$, $u\typ{\delta}$ where $\delta \in \Codata$, and $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$.
\end{itemize}
\item Repeat until $\mathcal{U}$ is unchanged
\begin{enumerate}
\item $\mathcal{U}' := \emptyset$
\item for each $U_i \in \mathcal{U}$:
\begin{itemize}
\item[\ ] Let $\mathcal{U}_i$ be a partition of $U_i$ such that $\ec{t} =_{\mathcal{U}_i} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if:
\begin{itemize}
\item $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$, and
\item $\ec{t_j} =_{\mathcal{U}} \ec{u_j}$ for each $j = 1, \ldots, k$.
\end{itemize}
\item[\ ] $\mathcal{U}' := \mathcal{U}' \cup \mathcal{U}_i$.
\end{itemize}
\item $\mathcal{U} := \mathcal{U}'$
\end{enumerate}
\item If distinct $\ec{t}, \ec{u} \in U$ for some $U \in \mathcal{U}$, apply \rn{Unique} to $t,u$.
\end{enumerate}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Algorithm for checking Uniqueness.
}
\label{fig:uniqueness-algo}
\end{figure}

\flushleft $\mathrm{traverse}( \ec{t}, U, P )$:
\begin{enumerate}
\item If $\ec{t} \in P$, apply \rn{Acyclic} to $t$.
\item If $\ec{t} \in U$,
\begin{enumerate}
\item If $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$, then for each $j = 1, \ldots k$,
\begin{itemize}
\item[\ ] $\mathrm{traverse}( \ec{t_j}, U, P \cup \{ \ec{t} \} )$
\end{itemize}
\item $U := U \textbackslash \{ \ec{t} \}$.
\end{enumerate}
\end{enumerate}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Algorithm for checking Acyclic.
}
\label{fig:acyclic-algo}
\end{figure}

@@

{\footnotesize
\def\ackname{Acknowledgment}
\paragraph{\footnotesize\ackname.}
\looseness=-1
We owe a great debt to the development team of CVC4,
including Clark Barrett and Cesare Tinelli,
and in particular Morgan Deters,
who jointly with the first author
developed the initial version
of the theory solver for datatypes in CVC4.
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatypes or at least benevolently tolerated it, both of which we are thankful
for.
%
Peter Gammie and Andreas Lochbihler provided useful benchmarks.
Andrei Popescu helped clarify our thoughts regarding codatatypes and indicated
related work. Dmitriy Traytel took part in discussions about degenerate
codatatypes.
%
Pascal Fontaine, Andreas Lochbihler, Andrei Popescu, Christophe Ringeissen, Mark
Summerfield, Dmitriy Traytel, and the anonymous reviewers suggested many
textual improvements. The second author's work was partially supported by
the Deutsche
Forschungs\-gemein\-schaft %(DFG)
project ``Den Hammer h\"arten'' (grant NI\,491\slash 14-1)
and the Inria technological development
action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC).

>>>>

\flushleft $\mathrm{traverse}( \ec{t}, U, P )$:
\begin{enumerate}
\item If $\ec{t} \in P$, apply \rn{Acyclic} to $t$.
\item If $\ec{t} \in U$,
\begin{enumerate}
\item If $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$, then for each $j = 1, \ldots k$,
\begin{itemize}
\item[\ ] $\mathrm{traverse}( \ec{t_j}, U, P \cup \{ \ec{t} \} )$
\end{itemize}
\item $U := U \textbackslash \{ \ec{t} \}$.
\end{enumerate}
\end{enumerate}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Algorithm for checking \rn{Acyclic}}
\label{fig:acyclic-algo}
\end{figure}

\begin{figure}[t!]
\normalsize
\centering
\begin{enumerate}
\item Let $\mathcal{U}$ be a partition of the equivalence classes of $\Ec$ such that $\ec{t} =_{\mathcal{U}} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if:
\begin{itemize}
\item $t\typ{\delta}$, $u\typ{\delta}$ where $\delta \in \Codata$, and $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$.
\end{itemize}
\item Repeat until $\mathcal{U}$ is unchanged
\begin{enumerate}
\item $\mathcal{U}' := \emptyset$
\item for each $U_i \in \mathcal{U}$:
\begin{itemize}
\item[\ ] Let $\mathcal{U}_i$ be a partition of $U_i$ such that $\ec{t} =_{\mathcal{U}_i} \ec{u}$ for distinct $\ec{t}, \ec{u}$ if and only if:
\begin{itemize}
\item $\const{C}( t_1, \ldots, t_k ) \in \ec{t}$ and $\const{C}( u_1, \ldots, u_k ) \in \ec{u}$, and
\item $\ec{t_j} =_{\mathcal{U}} \ec{u_j}$ for each $j = 1, \ldots, k$.
\end{itemize}
\item[\ ] $\mathcal{U}' := \mathcal{U}' \cup \mathcal{U}_i$.
\end{itemize}
\item $\mathcal{U} := \mathcal{U}'$
\end{enumerate}
\item If distinct $\ec{t}, \ec{u} \in U$ for some $U \in \mathcal{U}$, apply \rn{Unique} to $t,u$.
\end{enumerate}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Algorithm for checking \rn{Unique}}
\label{fig:uniqueness-algo}
\end{figure}

Before constructing a model for $F\!$, the theory solver constructs neither $\mu$-terms nor the mapping $\Val$.
Instead, we use the algorithms in Figures~\ref{fig:acyclic-algo} and~\ref{fig:uniqueness-algo}
for determining whether the rules \rn{Acyclic} and \rn{Unique} apply to our set $\Ec$.

For \rn{Acyclic}, Figure~\ref{fig:acyclic-algo} considers the set $U$ of all datatype equivalence classes of $\Ec$.
We choose an arbitrary $\ec{t}$ in $U$ and call the recursive subprocedure $\mathrm{traverse}$,
which takes as input the current equivalence class we are processing, the set of equivalence classes $U$ we have yet to process,
and the set of equivalence classes $P$ we are currently processing.
If a call to $\mathrm{traverse}( \ec{t}, U, P )$ is such that $\ec{t} \in P$,
we know that the rule \rn{Acyclic} applies to $t$.
Otherwise, if we have yet to process $\ec{t}$, as indicated by the condition $\ec{t} \in U$,
then if $\ec{t}$ contains a constructor term $\const{C}( t_1, \ldots, t_k )$, 
then we recursively call $\mathrm{traverse}$ on each of $\ec{t_1}, \ldots, \ec{t_n}$.
If this succeeds, we remove $\ec{t}$ from $U$.

For \rn{Unique}, Figure~\ref{fig:uniqueness-algo} considers a partition $\mathcal{U}$ of the equivalence classes of our set $\Ec$
such that equivalence classes having a constructor term with top symbol $\const{C}$ are placed in the same subset, for each constructor $\const{C}$.
We write $\ec{t} =_{\mathcal{U}} \ec{u}$ to denote that $\ec{t}$ and $\ec{u}$ reside in the same subset within $\mathcal{U}$.
We then refine $\mathcal{U}$ by 
constructing a partition $\mathcal{U}_i$ of each $U_i \in \mathcal{U}$ such that distinct $\ec{t}$ and $\ec{u}$ reside in the same subset of $\mathcal{U}_i$ 
if and only if they contain constructor terms $\const{C}( t_1, \ldots, t_k )$ and $\const{C}( u_1, \ldots, u_k )$
where each $t_j, u_j$ are such that $\ec{t_j}$ and $\ec{u_j}$ reside in the same subset of $\mathcal{U}$ for each $j=1, \ldots, k$.
Let $\mathcal{U'}$ be the union of each of these partitions.
We update $\mathcal{U}$ to $\mathcal{U'}$.
This refinement is repeated until $\mathcal{U}$ is left unchanged.
If the resulting partition $\mathcal{U}$ contains a subset $U$ having two distinct equivalence classes $\ec{t}$ and $\ec{u}$, then \rn{Unique} applies to $t$ and $u$.

%Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
%To detect whether \rn{Acyclic} applies,
%the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
%It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all constructor arguments in these equivalence classes recursively.
%If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
%To recognize when the precondition of \rn{Unique} holds,
%the procedure considers the set $\SSS$ of all codatatype equivalence classes. It
%simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
%while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
%and the equivalence of their arguments of ordinary types.
%It then partitions each set %in this partition %in the same manner
%recursively.
%If %after doing so,
%the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
%and $v$, it deduces that \rn{Unique} applies to $u$ and~$v$.

@@

@@

{
\def\ackname{Acknowledgment}
\paragraph{\ackname.}
\looseness=-1
We owe a great debt to the development team of CVC4,
including Clark Barrett and Cesare Tinelli,
and in particular Morgan Deters,
who jointly with the first author
developed the initial version
of the theory solver for datatypes in CVC4.
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatypes or at least benevolently tolerated it, both of which we are thankful
for.
%
Peter Gammie and Andreas Lochbihler provided useful benchmarks.
Andrei Popescu helped clarify our thoughts regarding codatatypes and indicated
related work. Dmitriy Traytel took part in discussions about degenerate
codatatypes.
%
Pascal Fontaine, Andreas Lochbihler, Andrei Popescu, Christophe Ringeissen, Mark
Summerfield, Dmitriy Traytel, and the anonymous reviewers suggested many
textual improvements. The second author's work was partially supported by
the Deutsche
Forschungs\-gemein\-schaft %(DFG)
project ``Den Hammer h\"arten'' (grant NI\,491\slash 14-1)
and the Inria technological development
action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC).
